---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# longcfa

<!-- badges: start -->
<!-- badges: end -->

The goal of longcfa is to simplify the specification of longitudinal confirmatory factor analysis (CFA) using the `lavaan` package using a matrix-like approach.

## Installation

You can install the development version of longcfa from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("marklhc/longcfa")
```

## Example

This is a basic example. See `vignette("longcfa")` for a longer example. The package also supports

- Invariance constraints and partial invariance, as discussed in [this article](articles/invariance.html).
- Categorical indicators, as discussed in `vignette("cat")`.
- Penalized estimation to obtain approximate invariance, as discussed in `vignette("penalized")`.
- Automatic specification search by relaxing invariance constraints one at a time, as discussed in `vignette("specification-search")`.

```{r example}
library(longcfa)
library(lavaan)
# Indicator matrix
spec <- matrix(
    c(
        "y1",
        "y2",
        "y3",
        "y4",
        "y5",
        "y6",
        "y7",
        "y8"
    ),
    ncol = 2
)
# Scalar invariance
fit <- longcfa(
    spec,
    lv_names = c("dem60", "dem65"),
    data = PoliticalDemocracy,
    long_equal = c("loadings", "intercepts")
)
summary(fit)
# Partial invariance
fit2 <- longcfa(
    spec,
    lv_names = c("dem60", "dem65"),
    data = PoliticalDemocracy,
    long_equal = c("loadings", "intercepts"),
    long_partial = list(
        loadings = matrix(c(1, 2), ncol = 2),
        intercepts = matrix(c(1, 3, 2, 2), ncol = 2)
    )
)
summary(fit2)
```


```{r}
# Approximate invariance with penalized unique covariances and differences on loadings and intercepts
# Starts with an unidentified model
mod3_un <- longcfa_syntax(
    spec,
    lv_names = c("dem60", "dem65"),
    lag_cov = TRUE,
    # Allow all unique covariances
    ucov_mat = t(combn(4, 2)),
    free_latvars = TRUE,
    free_latmeans = TRUE
)
lavaanify(mod3_un) # check parameter IDs for applying penalties
# Dry run
fit3 <- cfa(
    mod3_un,
    data = PoliticalDemocracy,
    std.lv = TRUE,
    do.fit = FALSE,
    start = fit2
)
# Penalized estimation
pen_fit3 <- penalized_est(
    fit3,
    w = .03,
    # penalize concurrent unique covariances
    pen_par_id = c(13:18, 31:36),
    pen_diff_id = list(
        loadings = rbind(1:4, 19:22),
        intercepts = rbind(5:8, 23:26)
    )
)
summary(pen_fit3)
```
