---
title: "Categorical Indicators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(longcfa)
library(lavaan)
```

## All Indicators Categorical

We first discretize the indicators of the `PoliticalDemocracy` dataset from the `lavaan` package. The first two indicators will be dichotomized, and the last two indicators will be trichotomized.

```{r}
# Discretized the PoliticalDemocracy data
data("PoliticalDemocracy", package = "lavaan")
pd_cat <- PoliticalDemocracy
# Dichotomize first two indicators
pd_cat[, c(1:2, 5:6)] <- lapply(PoliticalDemocracy[, c(1:2, 5:6)], function(x) {
    cut(x, breaks = c(-Inf, 5.0, Inf), labels = FALSE)
})
# Trichotomize the last two indicators
pd_cat[, c(3:4, 7:8)] <- lapply(PoliticalDemocracy[, c(3:4, 7:8)], function(x) {
    cut(x, breaks = c(-Inf, 2.5, 7.5, Inf), labels = FALSE)
})
```

### Configural model

Same factor structure across time points, no constraints on measurement parameters.

```{r}
# Using lavaan
cfa(
    "
dem60 =~ y1 + y2 + y3 + y4
dem65 =~ y5 + y6 + y7 + y8
",
    data = pd_cat,
    std.lv = TRUE,
    ordered = TRUE,
    parameterization = "theta"
) |>
    summary()
# Using longcfa
ind_mat <- matrix(
    c("y1", "y2", "y3", "y4",
      "y5", "y6", "y7", "y8"
    ),
    nrow = 4
)
longcfa(
    ind_mat,
    lv_names = c("dem60", "dem65"),
    data = pd_cat,
    ordered = TRUE
) |> 
    summary()
```

The warning message is likely due to the latent correlation being estimated to be larger than 1, which could be due to the arbitrary discretization of the indicators.

### Strict invariance

```{r}
longcfa(
    ind_mat,
    lv_names = c("dem60", "dem65"),
    data = pd_cat,
    ordered = TRUE,
    long_equal = c("loadings", "thresholds")
) |>
    summary()
```

## Mixed Continuous and Categorical

Here, only the first two indicators are categorical.

```{r}
pd_mixed <- pd_cat
pd_mixed[, c(1:2, 5:6)] <- PoliticalDemocracy[, c(1:2, 5:6)]
# Configural, using lavaan
cfa(
    "
dem60 =~ y1 + y2 + y3 + y4
dem65 =~ y5 + y6 + y7 + y8
",
    data = pd_mixed,
    std.lv = TRUE,
    ordered = paste0("y", c(3:4, 7:8)),
    parameterization = "theta"
) |>
    summary()
# configural, using longcfa
longcfa(
    ind_mat,
    lv_names = c("dem60", "dem65"),
    data = pd_mixed,
    ordered = c("y3", "y4", "y7", "y8")
)
```

```{r}
# strict invariance, using longcfa
longcfa(
    ind_mat,
    lv_names = c("dem60", "dem65"),
    data = pd_mixed,
    ordered = c("y3", "y4", "y7", "y8"),
    long_equal = c("loadings", "intercepts", "thresholds", "residuals")
) |>
    summary()
```