---
title: "Approximate Longitudinal Invariance With Alignment Optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Approximate Longitudinal Invariance With Alignment Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(longcfa)
library(lavaan)
library(sirt)
```

The data is the `PoliticalDemocracy` dataset from the `lavaan` package, which is a longitudinal dataset on political democracy scores across multiple countries over several years. One can read the [documentation](https://rdrr.io/cran/lavaan/man/PoliticalDemocracy.html) for more details.

```{r}
data("PoliticalDemocracy", package = "lavaan")
head(PoliticalDemocracy)
```

In this data set, `y1` to `y4` are indicators of political democracy in 1960, and `y5` to `y8` are the same indicators in 1965. Using `lavaan`, one can specify a longitudinal confirmatory factor analysis (CFA) model as follows:

```{r}
# Configural invariance: no constraints across time
lconfig_mod <- "
  dem60 =~ y1 + y2 + y3 + y4
  dem65 =~ y5 + y6 + y7 + y8
"
lconfig_fit <- cfa(
    lconfig_mod,
    data = PoliticalDemocracy,
    std.lv = TRUE,
    meanstructure = TRUE
)
lconfig_fit
```

```{r}
ind_mat <- matrix(
    c("y1", "y2", "y3", "y4", "y5", "y6", "y7", "y8"),
    nrow = 4
)
```

## Alignment optimization

The idea of alignment optimization ([Asparouhov & Muthén, 2014](https://www.statmodel.com/examples/webnotes/webnote18_3.pdf)) is to find, among all solutions that have the same fit as the configural invariance model, approximate measurement invariance by minimizing the amount of non-invariance in the model parameters. 

The package does not directly import functions from `sirt`, but instead contains some helper functions to help prepare inputs to pass to `sirt::invariance.alignment()`, as shown below.

```{r}
ld_mat <- get_lav_par_mat(lconfig_fit, op = "=~", ind_matrix = ind_mat)
int_mat <- get_lav_par_mat(lconfig_fit, op = "~1", ind_matrix = ind_mat)
aligned <- invariance.alignment(t(ld_mat), t(int_mat), meth = 2)
aligned_fit <- update_ustart(
    lconfig_fit,
    c(18, 29),
    c(aligned$pars$psi0[2]^2, aligned$pars$alpha0[2]),
    se = "none" # SE not trustworthy after alignment with arbitrary constraints
)
```

### Two-stage estimation

```{r}
fs_aligned <- lavPredict(aligned_fit, method = "Bartlett", acov = TRUE)
colnames(fs_aligned) <- paste0("fs_", colnames(fs_aligned))
attr(fs_aligned, "acov")
step2_fit <- cfa(
    "dem60 =~ 1 * fs_dem60
   dem65 =~ 1 * fs_dem65
   fs_dem60 ~~ 1.294956e-01 * fs_dem60
   fs_dem65 ~~ 1.084581e-01 * fs_dem65
   dem60 ~~ dem65",
    data = fs_aligned,
    meanstructure = TRUE
)
summary(step2_fit)
```
