---
title: "Specification Search for Partial Invariance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specification Search for Partial Invariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(longcfa)
library(lavaan)
```

This vignette demonstrates how to use the `plinv_search()` function to perform a specification search for partial invariance in longitudinal CFA models. We will use the `PoliticalDemocracy` dataset from the `lavaan` package.

## Data Preparation

First, we define the indicator matrix and latent variable names for the `PoliticalDemocracy` dataset.

```{r}
data("PoliticalDemocracy", package = "lavaan")

# Define indicator matrix (rows = indicators, columns = time points)
ind_mat <- cbind(
    c("y1", "y2", "y3", "y4"),
    c("y5", "y6", "y7", "y8")
)

# Define latent variable names
lv_names <- c("dem60", "dem65")
```

## Invariance Model

We start by fitting a model with strict longitudinal invariance constraints (on loadings, intercepts, and unique variances).

```{r}
fit_strict <- longcfa(
    ind_mat,
    lv_names = lv_names,
    data = PoliticalDemocracy,
    lag_cov = TRUE,
    long_equal = c("loadings", "intercepts", "residuals")
)

summary(fit_strict)
```

## Specification Search

The `plinv_search()` function iteratively relaxes invariance constraints based on modification indices or score tests. Here, we search for partial invariance in both loadings and intercepts.

We use `get_lav_test_score` as the "modification index" function (`mi_fun`), which uses the score test (`lavaan::lavTestScore()`) to identify constraints to release. We set a minimum threshold (`mi_min`) of 3.84 (corresponding to p < .05 for 1 df).

```{r}
get_lav_test_score(fit_strict, ind = paste0("y", 1:8), op = "~~")
```

Note that the `plinv_search()` function proceeds in steps, so when `type = c("loadings", "intercepts", "residuals")` is specified, it will first search for loadings from the metric invariance model, then intercepts, and finally residuals.

```{r}
# Perform specification search
search_res <- plinv_search(
    ind_matrix = ind_mat,
    lv_names = lv_names,
    data = PoliticalDemocracy,
    type = c("loadings", "intercepts", "residuals"),
    mi_fun = get_lav_test_score,
    mi_min = 3.84,
    lag_cov = TRUE # Passed to longcfa()
)
```

## Results

The result object contains the final fitted model (`fit`) and a trace of the relaxed constraints (`traces`).

### Relaxed Constraints

We can examine which constraints were relaxed during the search process.

```{r}
search_res$traces
```

The output shows the left-hand side (`lhs`), operator (`op`), and right-hand side (`rhs`) of the parameters that were freed, along with the modification index (`mi`) at that step.

### Final Model Summary

We can inspect the summary of the final partial invariance model.

```{r}
summary(search_res$fit)
```

### Compare Models

Finally, we can compare the initial strict invariance model with the partial invariance model found by the search.

```{r}
anova(fit_strict, search_res$fit)
```
