---
title: "Specification Search for Partial Invariance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specification Search for Partial Invariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(longcfa)
library(lavaan)
```

This vignette demonstrates how to use the `plinv_search()` function to perform a specification search for partial invariance in longitudinal CFA models. We will use the `PoliticalDemocracy` dataset from the `lavaan` package.

## Data Preparation

First, we define the indicator matrix and latent variable names for the `PoliticalDemocracy` dataset.

```{r}
data("PoliticalDemocracy", package = "lavaan")

# Define indicator matrix (rows = indicators, columns = time points)
ind_mat <- cbind(
    c("y1", "y2", "y3", "y4"),
    c("y5", "y6", "y7", "y8")
)

# Define latent variable names
lv_names <- c("dem60", "dem65")
```

## Invariance Model

We start by fitting a model with strict longitudinal invariance constraints (on loadings, intercepts, and unique variances).

```{r}
fit_strict <- longcfa(
    ind_mat,
    lv_names = lv_names,
    data = PoliticalDemocracy,
    lag_cov = TRUE,
    long_equal = c("loadings", "intercepts", "residuals")
)

summary(fit_strict)
```

## Specification Search

The `plinv_search()` function iteratively relaxes invariance constraints based on modification indices or score tests. Here, we search for partial invariance in both loadings and intercepts.

We use `get_lav_test_score` as the "modification index" function (`mi_fun`), which uses the score test (`lavaan::lavTestScore()`) to identify constraints to release. We set a minimum threshold (`mi_min`) of 3.84 (corresponding to p < .05 for 1 df).

```{r}
get_lav_test_score(fit_strict, ind = paste0("y", 1:8), op = "~~")
```

Note that the `plinv_search()` function proceeds in steps, so when `type = c("loadings", "intercepts", "residuals")` is specified, it will first search for loadings from the metric invariance model, then intercepts, and finally residuals.

```{r}
# Perform specification search
search_res <- plinv_search(
    ind_matrix = ind_mat,
    lv_names = lv_names,
    data = PoliticalDemocracy,
    type = c("loadings", "intercepts", "residuals"),
    mi_fun = get_lav_test_score,
    mi_min = 3.84,
    lag_cov = TRUE # Passed to longcfa()
)
```

## Results

The result object contains the final fitted model (`fit`) and a trace of the relaxed constraints (`traces`).

### Relaxed Constraints

We can examine which constraints were relaxed during the search process.

```{r}
search_res$traces
```

The output shows the left-hand side (`lhs`), operator (`op`), and right-hand side (`rhs`) of the parameters that were freed, along with the modification index (`mi`) at that step.

### Final Model Summary

We can inspect the summary of the final partial invariance model.

```{r}
summary(search_res$fit)
```

### Compare Models

Finally, we can compare the initial strict invariance model with the partial invariance model found by the search.

```{r}
anova(fit_strict, search_res$fit)
```

## Specification Search with Categorical Data

This section demonstrates how to perform a specification search when some or all indicators are categorical (ordinal). We discretize the continuous `PoliticalDemocracy` variables into ordinal categories and then constrain thresholds during invariance testing.

### Creating Ordered Indicators

We start by converting the first two items (`y1`, `y2`) into 3-category ordinal variables, and the remaining two items (`y3`, `y4`) into binary variables. The same transformation is applied at both time points.

```{r}
data("PoliticalDemocracy", package = "lavaan")

# Create ordered versions: 3 categories for y1/y2, 2 (binary) for y3/y4
ordinal_data <- PoliticalDemocracy

ordinal_data$y1_cat <- cut(
    ordinal_data$y1,
    breaks = 3,
    labels = c("low", "mid", "high")
)
ordinal_data$y2_cat <- cut(
    ordinal_data$y2,
    breaks = 3,
    labels = c("low", "mid", "high")
)

ordinal_data$y3_cat <- ifelse(
    PoliticalDemocracy$y3 >= median(PoliticalDemocracy$y3),
    "yes",
    "no"
)
ordinal_data$y4_cat <- ifelse(
    PoliticalDemocracy$y4 >= median(PoliticalDemocracy$y4),
    "yes",
    "no"
)

ordinal_data$y5_cat <- cut(
    ordinal_data$y5,
    breaks = 3,
    labels = c("low", "mid", "high")
)
ordinal_data$y6_cat <- cut(
    ordinal_data$y6,
    breaks = 3,
    labels = c("low", "mid", "high")
)

ordinal_data$y7_cat <- ifelse(
    PoliticalDemocracy$y7 >= median(PoliticalDemocracy$y7),
    "yes",
    "no"
)
ordinal_data$y8_cat <- ifelse(
    PoliticalDemocracy$y8 >= median(PoliticalDemocracy$y8),
    "yes",
    "no"
)

# Define indicator matrix for the ordinal items
ind_mat_ord <- cbind(
    c("y1_cat", "y2_cat", "y3_cat", "y4_cat"),
    c("y5_cat", "y6_cat", "y7_cat", "y8_cat")
)

# Specify which variables are ordered (all 8 items)
ordered_vars <- c(
    "y1_cat",
    "y2_cat",
    "y3_cat",
    "y4_cat",
    "y5_cat",
    "y6_cat",
    "y7_cat",
    "y8_cat"
)

# Define latent variable names
lv_names_ord <- c("dem60", "dem65")
```

### Fitting Strict Threshold Invariance Model

We fit a configural invariance model first, allowing thresholds to differ across time points. Then we test strict (scalar) threshold invariance by constraining corresponding thresholds equal across time.

```{r}
# Configural invariance: thresholds free across time
fit_conf <- longcfa(
    ind_mat_ord,
    lv_names = lv_names_ord,
    data = ordinal_data,
    ordered = ordered_vars,
    estimator = "WLSMV",
    lag_cov = TRUE,
    long_equal = c("loadings") # only constrain loadings
)

summary(fit_conf, fit.measures = TRUE)
```

### Specification Search on Thresholds

Now we use `plinv_search()` to find which threshold constraints should be relaxed. Support for threshold is currently experimental. We search only over thresholds (the `"thresholds"` type maps to the `"|"` operator internally). The function iteratively frees the threshold with the highest modification index until no remaining constraint exceeds `mi_min`.

```{r}
# Perform specification search on thresholds
search_thresh <- plinv_search(
    ind_matrix = ind_mat_ord,
    lv_names = lv_names_ord,
    data = ordinal_data,
    type = "thresholds",
    parameterization = "theta",
    mi_fun = get_lav_test_score,
    mi_min = 3.84,
    ordered = ordered_vars,
    estimator = "WLSMV",
    lag_cov = TRUE
)
```

### Relaxed Threshold Constraints

The search may free certain threshold parameters that show significant misspecification. We examine the trace of relaxed constraints:

```{r}
search_thresh$traces
```

The output shows which indicator and which threshold level was freed at each step, along with the modification index that justified the relaxation.

### Final Partial Threshold Invariance Model

We compare the configural model (all thresholds free) with the partial threshold invariance model found by the search:

```{r}
anova(fit_conf, search_thresh$fit)
summary(search_thresh$fit)
```

The partial threshold invariance model retains equality constraints on most thresholds while freeing only those that significantly violate scalar invariance. This provides a more accurate representation of measurement equivalence across time points for ordinal indicators.
