---
title: "A harmonization example with summary statistics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A harmonization example with summary statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following example demonstrates longitudinal harmonization using summary statistics. The data come from the summary statistics in [Tyrell et al. (2019)](https://doi.org/10.1080/15374416.2019.1622124), with two common parcels across five waves, but two unique parcels that are common in Waves 1-3, and two other unique parcels that are common in Waves 4-5. The correlation matrix, standard deviations, and means are provided in the paper, and imported below.

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(longcfa)

# Load the lavaan package
library(lavaan)

# 1. Define the variable names (20 variables total)
# fmt: skip
var_names <- c(
  "w1com1", "w1com2", "w1uniq1", "w1uniq2",
  "w2com1", "w2com2", "w2uniq1", "w2uniq2",
  "w3com1", "w3com2", "w3uniq1", "w3uniq2",
  "w4com1", "w4com2", "w4uniq1", "w4uniq2",
  "w5com1", "w5com2", "w5uniq1", "w5uniq2"
)

# 2. Input the lower triangular correlation matrix as a string
# We include the diagonal 1s to complete the matrix shape.
lower_cor_string <- "
  1
 .32 1
 .36 .45 1
 .20 .36 .35 1
 .27 .22 .28 .27 1
 .16 .31 .27 .23 .37 1
 .14 .22 .30 .18 .31 .50 1
 .20 .19 .24 .31 .44 .47 .38 1
 .15 .18 .24 .21 .25 .19 .22 .22 1
 .20 .30 .32 .10 .17 .30 .30 .22 .44 1
 .07 .19 .26 .13 .25 .27 .38 .28 .36 .43 1
 .17 .28 .21 .26 .22 .18 .14 .37 .32 .36 .25 1
 .00 .12 .11 .08 .16 .11 .10 .12 .23 .16 .26 .13 1
 .00 .16 .24 .14 .16 .07 .13 .04 .21 .19 .34 .15 .57 1
 .02 .20 .18 .13 .18 .11 .10 .16 .26 .18 .29 .21 .64 .69 1
-.06 .16 .19 .19 .13 .15 .07 .13 .17 .18 .25 .23 .50 .56 .67 1
 .07 .08 .15 .08 .13 .11 .19 .21 .19 .19 .31 .13 .31 .24 .31 .30 1
 .04 .13 .18 .12 .10 .17 .23 .14 .17 .17 .35 .09 .33 .45 .39 .36 .57 1
 .04 .19 .16 .16 .10 .08 .14 .10 .23 .19 .22 .15 .43 .44 .60 .45 .51 .61 1
 .05 .11 .14 .16 .11 .12 .13 .15 .14 .11 .20 .13 .33 .34 .39 .43 .40 .56 .53 1
"

# Parse the string into a full symmetric correlation matrix
cor_matrix <- lav_getcov(lower_cor_string, names = var_names)

# 3. Input Standard Deviations and Means
# fmt: skip
sds <- c(0.38, 0.49, 0.44, 0.50, 0.39, 0.44, 0.44, 0.47, 0.34, 0.44, 
         0.43, 0.50, 0.33, 0.40, 0.33, 0.29, 0.33, 0.40, 0.28, 0.27)

# fmt: skip
means <- c(1.44, 1.42, 1.42, 1.41, 1.39, 1.40, 1.35, 1.42, 1.42, 1.41, 
           1.42, 1.51, 1.21, 1.29, 1.22, 1.18, 1.25, 1.32, 1.19, 1.16)

# Assign variable names to the vectors for safe keeping
names(sds) <- var_names
names(means) <- var_names

# 4. Convert Correlation matrix to Covariance matrix
cov_matrix <- lav_cor2cov(cor_matrix, sds)
```

```{r}
ind_mat <- matrix(NA, nrow = 6, ncol = 5)
ind_mat[1:4, 1:3] <- var_names[1:12]
ind_mat[c(1:2, 5:6), 4:5] <- var_names[13:20]
ind_mat
```

```{r}
lconfig_fit <- longcfa(
    ind_mat,
    lv_names = paste0("w", 1:5, "dep"),
    sample.cov = cov_matrix,
    sample.mean = means,
    sample.nobs = 392,
    lag_cov = TRUE
)
summary(lconfig_fit, fit.measures = TRUE)
```

```{r}
pen_fit <- penalized_longcfa(
    ind_mat,
    lv_names = paste0("w", 1:5, "dep"),
    sample.cov = cov_matrix,
    sample.mean = means,
    sample.nobs = 392,
    lag_cov = TRUE,
    pen_fn = "l0a",
    w = 0.1
)
```


```{r}
summary(pen_fit)
# Penalized estimates of loadings and intercepts
(load_mat <- longcfa::get_lav_par_mat(pen_fit, "=~", ind_matrix = ind_mat))
(int_mat <- longcfa::get_lav_par_mat(pen_fit, "~1", ind_matrix = ind_mat))
# Effective number of loadings
eff_load_diff <- plavaan::composite_pair_loss(load_mat, fun = plavaan::l0a)
cat("Effective number of non-invariant loadings:", eff_load_diff, "\n")
# Effective number of intercepts
eff_int_diff <- plavaan::composite_pair_loss(int_mat, fun = plavaan::l0a)
cat("Effective number of non-invariant intercepts:", eff_int_diff, "\n")
```
